### 2017-05-26 Paper_v1.tex
May26

The reaction rate was written as the product of a kinetic term ($\bar{r}_{j}$) and a control term ($v_{j}$), $r_{j}\left(\mathbf{x},\mathbf{k}\right)=\bar{r}_{j}v_{j}$.
We used multiple saturation kinetics to model the reaction term $\bar{r}_{j}$:
\begin{equation}\label{eqn:rate-bar}
	\bar{r}_{j}=V_{j}^{max}\epsilon_{i}\prod_{s\in{m_{j}^{-}}}\frac{x_{s}}{K_{js} + x_{s}}
\end{equation}
where $V_{j}^{max}$ denotes the maximum rate for reaction $j$, $\epsilon_{i}$ denotes the scaled enzyme activity which catalyzes reaction $j$, $K_{js}$ denotes the saturation constant for species $s$, in reaction $j$ and $m_{j}^{-}$ denotes the set of \textit{reactants} for reaction $j$.
On the other hand, the control term $0\leq v_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
For each rate, we used a rule-based approach to select from competing control factors.
If rate j was influenced by $1,\dots,m$ factors, we modeled this relationship as
$v_{j}=\mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a transfer function quantifying the influence of factor $i$ on rate $j$.
The function $\mathcal{I}_{j}\left(\cdot\right)$ is an integration rule which maps the output of regulatory transfer functions into a control
variable. We used hill-like transfer functions and $\mathcal{I}_{j}\in\left\{min,max\right\}$ in this study \cite{pr3010138}.

@@

Transcription was modeled as first-order in RNA polymerase, since it acts like an enzyme:
\begin{equation}\label{eqn:transcription}
    \bar{r}^{TX} = V_{max}^{TX}\cdot{\rm RNAP}\left(\frac{{\rm GENE}}{K_{GENE}^{TX}+{\rm GENE}}\right)\epsilon_{i}\prod_{s\in{m_{TX}^{-}}}\frac{x_{s}}{K_{s}^{TX}+x_{s}}
\end{equation}
where $V_{max}^{TX}$ denotes the maximum transcription rate, ${\rm RNAP}$ denotes the RNA polymerase concentration, $GENE$ denotes the gene concentration, $K_{GENE}^{TX}$ denotes the gene saturation constant, $\epsilon_{i}$ denotes the scaled enzyme activity, $K_{s}^{TX}$ denotes the saturation constant for species $s$, and $m_{TX}^{-}$ denotes the set of \textit{reactants} for transcription: ATP, GTP, CTP, UTP, and water.
While transcription is modeled as saturating with respect to gene concentration, the gene is not considered a reactant in the stoichiometry as it is not consumed.

Translation was modeled as first-order in ribosome, since it acts like an enzyme:
\begin{equation}\label{eqn:translation}
    \bar{r}^{TL} = V_{max}^{TL}\cdot{\rm RIBO}\left(\frac{{\rm mRNA}}{K_{mRNA}^{TL}+{\rm mRNA}}\right)\epsilon_{i}\prod_{s\in{m_{TL}^{-}}}\frac{x_{s}}{K_{s}^{TL}+x_{s}}
\end{equation}
where $V_{max}^{TL}$ denotes the maximum translation rate, ${\rm RIBO}$ denotes the ribosome concentration, ${\rm mRNA}$ denotes the transcript concentration, $K_{mRNA}^{TL}$ denotes the transcript saturation constant, $\epsilon_{i}$ denotes the scaled enzyme activity, $K_{s}^{TL}$ denotes the saturation constant for species $s$, and $m_{TL}^{-}$ denotes the set of \textit{reactants} for translation: GTP, water, and the 20 species representing tRNA charged with amino acids.
While translation is modeled as saturating with respect to transcript concentration, the transcript is not considered a reactant in the stoichiometry as it is not consumed.

Transcript degradation was modeled as first-order in transcript:
\begin{equation}\label{eqn:mRNA_degradation}
    \bar{r}_{deg}^{mRNA} = k_{deg}^{mRNA}\cdot{\rm mRNA}
\end{equation}
where $k_{deg}^{mRNA}$ denotes the transcript degradation rate constant.

@@

Model parameters were constrained by literature \cite{Milo:2009aa}.
The rate maximum for CAT transcription was calculated as 123 h$^{-1}$:
\begin{equation}\label{eqn:transcription}
	V_{max}^{TX}=\left(\frac{k_{TX}}{l_{mRNA}}\right)\cdot P
\end{equation}
where $k_{TX}$ denotes the average mRNA elongation rate (25 nt/s, \cite{Garamella:2016aa}), $l_{mRNA}$ denotes the CAT mRNA length (660 nt), and $P$ denotes the level of promoter activity (estimated at 0.9).
% $RNAP$ denotes the RNA polymerase concentration (67.5 nM, \cite{Garamella:2016aa}),
The rate maximum for CAT translation was calculated as 247 h$^{-1}$:
\begin{equation}\label{eqn:transcription}
	V_{max}^{TL}=\left(\frac{k_{TL}}{l_{protein}}\right)\cdot K_P
\end{equation}
where $k_{TL}$ denotes the peptide chain elongation rate (1.5 aa/s, \cite{Garamella:2016aa}), $l_{protein}$ denotes the CAT protein length (219 aa), and $K_P$ denotes the polysome amplification constant (estimated at 10).
% $Ribo$ denotes the ribosome concentration (2000 nM, \cite{Garamella:2016aa}),
The transcript degradation rate constant was calculated as 5.2 h$^{-1}$:
\begin{equation}\label{eqn:transcription}
	k_{deg}^{mRNA}=\left(\frac{log(2)}{t_{\sfrac{1}{2}}}\right)\cdot{\rm mRNA}
\end{equation}
where $t_{\sfrac{1}{2}}$ denotes the mRNA degradation time (8 min, BNID 106253).
%The rate maximum for tRNA charging was calculated as 4200 mM/h:
%\begin{equation}\label{eqn:transcription}
%	V_{max}^{tRNA}=k_{tRNA}\cdot{\rm tRNA}
%\end{equation}
%where $k_{tRNA}$ denotes the tRNA charging catalytic rate (39 s$^{-1}$, BNID 104980), and ${\rm tRNA}$ denotes the charged tRNA concentration (estimated at 30 $\mu$M, consistent with the kinetic model).
Transcription, translation, mRNA degradation, and tRNA charging were bounded within two orders of magnitude of their calculated values.
A characteristic cell-free enzyme concentration of 167 nM was calculated by diluting the one-tenth maximal amount \textit{lacZ} (5 $\mu$M, BNID 100735) by a cell-free dilution factor of 30.
This enzyme level was then used to calculate rate maxima from turnover numbers for various enzymes from literature (Table~\ref{tbl:Rate_maxima}).
Rate maxima were bounded within one order of magnitude of the calculated value where available; all other rate maxima were bounded within two orders of magnitude of the geometric mean of the available values.

>>>>

The reaction rate was written as the product of a kinetic term ($\bar{r}_{j}$) and a control term ($v_{j}$), $r_{j}\left(\mathbf{x},\mathbf{k}\right)=\bar{r}_{j}v_{j}$.
We used multiple saturation kinetics to model the reaction term $\bar{r}_{j}$:
\begin{equation}\label{eqn:rate-bar}
	\bar{r}_{j}=V_{j}^{max}\epsilon_{i}\prod_{s\in{m_{j}^{-}}}\frac{x_{s}}{K_{js} + x_{s}}
\end{equation}
where $V_{j}^{max}$ denotes the maximum rate for reaction $j$, $\epsilon_{i}$ denotes the scaled enzyme activity which catalyzes reaction $j$, $K_{js}$ denotes the saturation constant for species $s$, in reaction $j$ and $m_{j}^{-}$ denotes the set of \textit{reactants} for reaction $j$.
Transcription was modeled as first-order in RNA polymerase, since it acts like an enzyme:
\begin{equation}\label{eqn:transcription}
    \bar{r}^{TX} = k_{cat}^{TX}\cdot{\rm RNAP}\left(\frac{{\rm GENE}}{K_{GENE}^{TX}+{\rm GENE}}\right)\prod_{s\in{m_{TX}^{-}}}\frac{x_{s}}{K_{s}^{TX}+x_{s}}
\end{equation}
where $k_{cat}^{TX}$ denotes the maximum transcription rate, ${\rm RNAP}$ denotes the RNA polymerase concentration, $GENE$ denotes the gene concentration, $K_{GENE}^{TX}$ denotes the gene saturation constant, $K_{s}^{TX}$ denotes the saturation constant for species $s$, and $m_{TX}^{-}$ denotes the set of \textit{reactants} for transcription: ATP, GTP, CTP, UTP, and water.
While transcription is modeled as saturating with respect to gene concentration, the gene is not considered a reactant in the stoichiometry as it is not consumed.
Translation was modeled as first-order in ribosome, since it acts like an enzyme:
\begin{equation}\label{eqn:translation}
    \bar{r}^{TL} = k_{cat}^{TL}\cdot{\rm RIBO}\left(\frac{{\rm mRNA}}{K_{mRNA}^{TL}+{\rm mRNA}}\right)\prod_{s\in{m_{TL}^{-}}}\frac{x_{s}}{K_{s}^{TL}+x_{s}}
\end{equation}
where $k_{cat}^{TL}$ denotes the maximum translation rate, ${\rm RIBO}$ denotes the ribosome concentration, ${\rm mRNA}$ denotes the transcript concentration, $K_{mRNA}^{TL}$ denotes the transcript saturation constant, $K_{s}^{TL}$ denotes the saturation constant for species $s$, and $m_{TL}^{-}$ denotes the set of \textit{reactants} for translation: GTP, water, and the 20 species representing tRNA charged with amino acids.
While translation is modeled as saturating with respect to transcript concentration, the transcript is not considered a reactant in the stoichiometry as it is not consumed.
Transcript degradation was modeled as first-order in transcript:
\begin{equation}\label{eqn:mRNA_degradation}
    \bar{r}_{deg}^{mRNA} = k_{deg}^{mRNA}\cdot{\rm mRNA}
\end{equation}
where $k_{deg}^{mRNA}$ denotes the transcript degradation rate constant.

The control term $0\leq v_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
For each rate, we used a rule-based approach to select from competing control factors.
If rate j was influenced by $1,\dots,m$ factors, we modeled this relationship as
$v_{j}=\mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a transfer function quantifying the influence of factor $i$ on rate $j$.
The function $\mathcal{I}_{j}\left(\cdot\right)$ is an integration rule which maps the output of regulatory transfer functions into a control
variable. We used hill-like transfer functions and $\mathcal{I}_{j}\in\left\{min,max\right\}$ in this study \cite{pr3010138}.

@@

@@

Model parameters were constrained by literature \cite{Milo:2009aa}.
The rate maximum for CAT transcription was calculated as 123 h$^{-1}$:
\begin{equation}\label{eqn:transcription}
	k_{cat}^{TX}=\left(\frac{k_{TX}}{l_{mRNA}}\right)\cdot P
\end{equation}
where $k_{TX}$ denotes the average mRNA elongation rate (25 nt/s, \cite{Garamella:2016aa}), $l_{mRNA}$ denotes the CAT mRNA length (660 nt), and $P$ denotes the level of promoter activity (estimated at 0.9).
% $RNAP$ denotes the RNA polymerase concentration (67.5 nM, \cite{Garamella:2016aa}),
The rate maximum for CAT translation was calculated as 247 h$^{-1}$:
\begin{equation}\label{eqn:transcription}
	k_{cat}^{TL}=\left(\frac{k_{TL}}{l_{protein}}\right)\cdot K_P
\end{equation}
where $k_{TL}$ denotes the peptide chain elongation rate (1.5 aa/s, \cite{Garamella:2016aa}), $l_{protein}$ denotes the CAT protein length (219 aa), and $K_P$ denotes the polysome amplification constant (estimated at 10).
% $Ribo$ denotes the ribosome concentration (2000 nM, \cite{Garamella:2016aa}),
The transcript degradation rate constant was calculated as 5.2 h$^{-1}$:
\begin{equation}\label{eqn:transcription}
	k_{deg}^{mRNA}=\left(\frac{log(2)}{t_{\sfrac{1}{2}}}\right)\cdot{\rm mRNA}
\end{equation}
where $t_{\sfrac{1}{2}}$ denotes the mRNA degradation time (8 min, BNID 106253).
%The rate maximum for tRNA charging was calculated as 4200 mM/h:
%\begin{equation}\label{eqn:transcription}
%	V_{max}^{tRNA}=k_{tRNA}\cdot{\rm tRNA}
%\end{equation}
%where $k_{tRNA}$ denotes the tRNA charging catalytic rate (39 s$^{-1}$, BNID 104980), and ${\rm tRNA}$ denotes the charged tRNA concentration (estimated at 30 $\mu$M, consistent with the kinetic model).
Transcription, translation, mRNA degradation, and tRNA charging were bounded within two orders of magnitude of their calculated values.
A characteristic cell-free enzyme concentration of 167 nM was calculated by diluting the one-tenth maximal amount \textit{lacZ} (5 $\mu$M, BNID 100735) by a cell-free dilution factor of 30.
This enzyme level was then used to calculate rate maxima from turnover numbers for various enzymes from literature (Table~\ref{tbl:Rate_maxima}).
Rate maxima were bounded within one order of magnitude of the calculated value where available; all other rate maxima were bounded within two orders of magnitude of the geometric mean of the available values.
